---
title: "Week 4 Lab, MAST20005/MAST90058"
author: "School of Mathematics and Statistics, University of Melbourne"
date: "Semester 2, 2022"
institute: "University of Melbourne"
fontsize: 11pt
classoption: a4paper
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
---

**Goals:** (i) Compute and interpret confidence intervals; (ii) Assess
properties of estimators using numerical simulations.

**Data for the [Confidence intervals] section:** Weight of primary tumour in
mice (`PTweight.txt`).  Observations represent measurements on primary tumour
weight (in micro-grams) in mice randomly assigned to stress and control groups.
Measurements are taken after a follow-up period. The goal of the study is to
compare tumour development in the two groups of mice.  (Data source: Sloan
Biology Lab, Monash University)

**Data for the [Comparing estimators by simulation] section:** Ivadopodia data
(`invadopodia.txt`). Invadopodia are protrusions in cells which are often
associated with cancer invasiveness and metastasis. The first dataset consists
of counts for the number of cells susceptible of invadopodia development from
mice tissue samples. The first column (`Condition`) denotes treatment group (1
= no treatment, 2 = isopropile (iso), 3 = propile (pro), 4 = iso + pro), while
the second column (`Total`) denotes cell counts.  (Data source: Sloan Biology
Lab, Monash University)

```{r echo=FALSE}
set.seed(2253)
```

# Confidence intervals

While constructing confidence intervals, we will often use R to find quantiles
of common distributions, including $\chi^2$-, $t$- and $F$-distributions.  For
each of the following distributions, find the 0.9, 0.95 and 0.975 quantiles.

1. Standard normal, $\mathrm{N}(0, 1)$
```{r }
p <- c(0.9, 0.95, 0.975)
qnorm(p)
```

2. $\mathrm{N}(5, 3^2)$
```{r eval=FALSE}
qnorm(p, 5, 3)
```

3. $t_5$
```{r eval=FALSE}
qt(p, 5)
```

4. $\chi_1^2$
```{r eval=FALSE}
qchisq(p, 1)
```

5. $\chi_5^2$
```{r eval=FALSE}
# your turn...
```

6. $\mathrm{F}_{12, 4}$
```{r eval=FALSE}
qf(p, 12, 4)
```

Load the primary tumour data into a data frame called `PTweight` (how?). Find
an approximate 90\% confidence interval for the mean tumour weight in the
control group.
```{r echo=FALSE}
PTweight <- read.table("PTweight.txt")
```
```{r }
x <- PTweight[1:5, 2]
n <- length(x)
x.bar <- mean(x)
s <- sd(x)
t <- qt(0.95, n - 1)
x.bar + c(-1, 1) * t * s / sqrt(n)
```
Notice that in the last line the vector $(-1,1)$ is multiplied and added to
scalars.  This produces a two-element vector corresponding to $\bar{x} \pm
t_{n-1}^{-1}(0.05) s / \sqrt{n}$.  Next compare your result with the in-built
function `t.test()`.
```{r }
t.test(x, conf.level = 0.90)
```
The last command is what we use in practice. We will look at the hypothesis
testing part of the output later in the course.

Construct a 95\% confidence interval for the difference of means $\mu_X
- \mu_Y$ in the case and control groups. Can you conclude that stress and
tumour growth are associated?  (Assume equal variances in the two groups,
i.e.\ $\sigma^2_X = \sigma_Y^2$)
```{r }
y <- PTweight[6:10, 2]  # stress group data
y.bar <- mean(y)
s.p <- sqrt((4 * var(x) + 4 * var(y)) / 8)  # pooled sample sd
x.bar - y.bar + c(-1, 1) * qt(0.975, df = 8) * s.p * sqrt(1 / 5 + 1 / 5)
```
The value $\mu_X - \mu_Y <- 0$ is inside the interval. Thus, there is not
enough evidence in these data to claim that the  means in the two groups are
different.

Compare now with the result from `t.test()`.
```{r }
t.test(x, y, var.equal = TRUE)
```

Is it reasonable to assume that the two variances are the same in the two
groups? This can be checked formally by computing a confidence interval for the
variance ratio $\sigma^2_X/ \sigma^2_Y$  as follows:
```{r }
var.test(x, y)
```
Since the value $\sigma^2_X/ \sigma^2_Y=1$ is inside the CI, there is not
enough evidence to suggest that the variances are unequal.


# Comparing estimators by simulation

Let $X_1,\dots,X_n$ be a random sample of size $n$ from a Poisson distribution
with mean $\lambda$. Let $\bar{X}$ and $S^2$ denote the sample mean and sample
variance, respectively. Note that both are unbiased estimators for $\lambda$
since
\[ \mathbb{E}(\bar{X}) = \mathbb{E}(S^2) = \lambda. \]

While we already know that $\mathrm{var}(\bar{X}) = \lambda/n$ (recall how), an
expression for $\mathrm{var}(S^2)$ is harder to compute. In such situations
computer simulations can help us compare estimators. First let us check that
the two estimators are unbiased.
```{r }
lambda <- 10
B <- 1000  # simulation runs
n <- 10    # sample size
xbar <- 1:B  # we will collect results in these two vectors
s2   <- 1:B
for (b in 1:B) {  # repeat B times for b = 1,...,B
    x <- rpois(n, lambda)
    xbar[b] <- mean(x)  # compute and store X-bar
    s2[b]   <- var(x)   # compute and store S^2
}
```
The LLN for iid variables $Z_1,\dots,Z_B$ states that $B^{-1} \sum_{b=1}^B Z_b
\rightarrow \mathbb{E}(Z_1)$ as $B \rightarrow \infty$. This can be used to
approximate the $\mathbb{E}(\bar{X})$ and $\mathbb{E}(S^2)$ from our simulations
as follows:
```{r }
mean(xbar)
mean(s2)
```
Both estimators seem to be unbiased for $\lambda$.

Using the LLN we can also approximate the variance of the two estimators as
follows:
```{r }
var(xbar)
var(s2)
```
While we already know that $\mathrm{var}(\bar{X}) = \lambda/n$, a more involved
calculation shows that $\mathrm{var}(S^2) = [\lambda(2n\lambda + n - 1)] /
[{n (n - 1)}]$.  The values obtained from the above simulation are quite close
to these theoretical values:
```{r }
lambda / n
lambda * (2 * n * lambda + n - 1) / (n * (n - 1))
```
Both estimators seem to be unbiased for $\lambda$ but the first estimator is
clearly superior in terms of variance. Try different values of the true
parameter $\lambda$ and $n$ in the above code and see what happens.  Also,
check what happens when you increase $B$.

Compare the accuracy of the estimators graphically, for example using boxplots.
```{r fig.height=4, fig.width=6}
boxplot(xbar, s2, names = c("X-bar", "S-squared"),
        col = "lightblue", horizontal = TRUE)

# Draw a dashed vertical line showing the true parameter value.
abline(v = lambda, lty = 2, lwd = 2, col = "magenta")
```

From the box plots, the distribution of both estimators is clearly centred
around the true $\lambda$, but the distribution of $\bar{X}$ has a much smaller
spread. Note also that the distribution of $S^2$ is not symmetric.

Load the invadopodia data. Assume a Poisson model with mean $\lambda_j$, $j
= 1, \dots, 4$ with different means corresponding to treatment groups and
estimate $\lambda_j$. For the first two groups we have:
```{r }
invadopodia <- read.table("invadopodia.txt")
x1 <- invadopodia[invadopodia$Condition == 1, 2]
x2 <- invadopodia[invadopodia$Condition == 2, 2]
x.bar1 <- mean(x1)
x.bar2 <- mean(x2)
```

Construct approximate 95\% confidence intervals for $\lambda$ recalling that
$\hat{\lambda} = \bar{X}$ obeys the Central Limit Theorem.  Specifically,
$\hat{\lambda} \approx \mathrm{N}(\lambda, \lambda/n)$, so approximate 95\%
confidence intervals can be computed using:
\[ \hat{\lambda}_j \pm 1.96 \times \sqrt{\frac{\hat{\lambda}_j}{n}}, \quad
   j = 1, \dots, 4. \]
```{r }
# 95% CI for group 1.
x.bar1 + c(-1, 1) * 1.96 * sqrt(x.bar1 / length(x1))
# 95% CI for group 2.
x.bar2 + c(-1, 1) * 1.96 * sqrt(x.bar2 / length(x2))
```
Note that the 1.96 in the above code gives 95\% confidence intervals. To
compute 90\% or 99\% CIs, replace 1.96 by the appropriate standard normal
quantiles:
```{r }
qnorm(0.95)   # use this to obtain a 90% CI
qnorm(0.995)  # use this to obtain a 99% CI
```

The CIs do not overlap, suggesting that the two means are actually different.
To carry out proper inference for the difference of means, $\lambda_1
- \lambda_2$, we need to derive a specific CI.  Since the first two groups are
independent we have,
\[ \hat{\lambda}_1 - \hat{\lambda}_2 \approx 
   \mathrm{N}\left(\lambda_1 - \lambda_2,
           \frac{\lambda_1 }{n_1} + \frac{\lambda_2}{n_2}\right) \]

```{r }
# 95% CI for difference between groups 1 and 2.
x.bar1 - x.bar2 + c(-1, 1) * 1.96 *
    sqrt(x.bar1 / length(x1) + x.bar2 / length(x2))
```

The interval is clearly above the value $\lambda_1 - \lambda_2 = 0$, suggesting
that the true difference is likely to be different from zero.  Hence, we
conclude that we have evidence that the isopropile treatment reduces the number
of susceptible cells in mice tissues.


# Simulating discrete distributions

The function `sample()` carries out sampling from a discrete
distribution in a few different ways.  For example, to simulate 10 coin tosses
you would use:
```{r }
sample(c("tails", "heads"), 10, replace = TRUE)
```
The first argument specifies the possible values to observe.  For example, we
could pick a random day of the week using:
```{r }
days <- c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")
sample(days, 1)
```
By default, all items are given equal probability.  This can be changed by
specifying different values using the `prob` argument.  For example,
biased coin tosses can be obtained with:
```{r }
sample(c("tails", "heads"), 10, prob = c(0.2, 0.8), replace = TRUE)
```
The vector passed to `prob` should have the same length as the set of
possible values (the first argument).

The argument `replace` specifies whether to do sampling with or without
replacement^['Replacement' refers to putting the sampled item back into the
pool of possible values before taking the next sample.].  The default is
without replacement, which means we always need to write `replace = TRUE` to
get iid samples.


# Exercises

1. Give estimates and 95\% CIs for $\lambda_3$ and $\lambda_4$.
```{r }
# Insert answer here...
```

2. The investigators are wondering whether isopropile treatment affects the
number of susceptible cells in mice tissues when propile treatment is already
provided.  What analysis answers that question?  Carry out this analysis.
```{r }
# Insert answer here...
```

3. Consider question 1 from the tutorial problems.  Derive a 75\% CI for the
population mean.
```{r }
# Insert answer here...
```

4. Consider question 2 from the week 3 tutorial problems.  This involved
a random sample of $n$ observations on $X$ having the following pmf:

    | $x$    | 0            | 1              | 2            |
    |:------:|:------------:|:--------------:|:------------:|
    | $p(x)$ | $1 - \theta$ | $3 \theta / 4$ | $\theta / 4$ |

    For the case of $n = 10$ and $\theta = 0.6$, use numerical simulations to
    show:

    a) $T_1 = (4 / 5) \bar{X}$ and $T_2 = 1- n^{-1} \sum_{i=1}^n I(X_i = 0)$
    are unbiased
    ```{r }
    # Insert answer here...
    ```

    b) $\mathrm{var}(T_1) > \mathrm{var}(T_2)$
    ```{r }
    # Insert answer here...
    ```
