---
title: "Week 7 Lab, MAST20005/MAST90058"
author: "School of Mathematics and Statistics, University of Melbourne"
date: "Semester 2, 2022"
institute: "University of Melbourne"
fontsize: 11pt
classoption: a4paper
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
---

**Goals:** (i) Basic hypothesis testing for proportions; (ii) Type I and type
II error; (iii) Power.

**Data:** 20 students counted the number of yellow lollies and the total number
of lollies in a 48.1 gram packet.  The data file, `candies.txt`, can be
obtained from the shared folder in the computer labs or from the LMS.

```{r echo=FALSE}
set.seed(2312)
```

# Type I and Type II errors

Let $X \sim \mathrm{Bi}(1, p)$ and let $X_1, \dots, X_{10}$ be a random sample
of size 10.  Consider a test of $H_0\colon p = 0.5$ against $H_1\colon
p = 0.25$. Let $Y = \sum_{i=1}^{10} X_i$. Define the critical region as $y
< 3.5$.

Find the value of $\alpha$, the probability of a Type I error. Do not use
a normal approximation (use the function `pbinom()`).
```{r }
pbinom(3.5, 10, 0.5)
pbinom(3, 10, 0.5)  # same answer as above!  (why?)
```

Find the value of $\beta$, the probability of a Type II error. Do not use
a normal approximation.
```{r }
1 - pbinom(3.5, 10, 0.25)
```

Next, carry out a simulation to estimate the Type I error of the test. Simulate
200 observations on $Y$ when $p = 0.5$ and then find the proportion of cases
when $H_0$ was rejected. Is this close to $\alpha$?
```{r }
T <- rbinom(200, 10, 0.5)  # simulate under H0
alpha <- sum(T < 3.5) / length(T)  # estimate
alpha
```

Simulate 200 observations on $Y$ when $p = 0.25$. Find the proportion of cases
when $H_0$ was not rejected. Is this close to $\beta$?
```{r }
T1 <- rbinom(200, 10, 0.25)  # simulate under H1
beta <- sum(T1 > 3.5) / length(T1) # estimate
beta
```

Estimates from simulations are affected by sampling error. The last two
questions may be answered more rigorously by computing confidence intervals for
your simulation-based estimates in the usual way (note that estimated
probabilities for $\alpha$ and $\beta$ are just sample proportions):
```{r }
alpha + c(-1 ,1) * 1.96 * sqrt(alpha * (1 - alpha) / 200) #  CI for alpha
beta  + c(-1, 1) * 1.96 * sqrt(beta  * (1 - beta)  / 200) #  CI for beta
```


# Power

Let $p$ be the probability that a tennis player's first serve is good. The
player takes lessons to increase $p$. After the lessons she wishes to test the
null hypothesis $H_0\colon p = 0.4$ against the alternative $H_1\colon p >
0.4$.  Let $y$ be the number out of $n = 25$ serves that are good, and let the
critical region be defined by $y \geqslant 13$.

Let the power function be $K(p) = \Pr(Y \geqslant 13 \mid p)$.  Graph
this function for $0 < p < 1$.
```{r fig.width=6, fig.height=4}
K1 <- function(p)
    1 - pbinom(12, 25, p)
p <- seq(0, 1, 0.01)
K <- K1(p)
plot(p, K, type = "l", ylab = "Power, K(p)")
```

Find the value of $\alpha = K(0.4)$.
```{r }
K1(0.4)
```

Find the value of $\beta$ when $p = 0.6$, ($\beta = 1 - K(0.6)$)
```{r }
1 - K1(0.6)
```

What happens to power when the sample size increases?  Suppose the player
carries out $n = 30$ serves:
```{r fig.width=6, fig.height=4}
K2 <- function(p)
    1 - pbinom(12, 30, p)
curve(K1, from = 0, to = 1, xlab = "p", ylab = expression(1 - beta))
curve(K2, from = 0, to = 1, add = TRUE, col = 2, lty = 2)
```


# Lollies data

Let $p$ be the proportion of yellow lollies in a packet of mixed colours. It is
claimed that $p = 0.2$.

Let's define a test statistic and an approximate critical region with a
significance level of $\alpha = 0.05$ to test $H_0\colon p = 0.2$ against
$H_1\colon p \ne 0.2$.

We reject $H_0$ if:
\[ \lvert z \rvert = \frac{\lvert \hat p - 0.2 \rvert}
                          {\sqrt{0.2 \times 0.8 / n}} > 1.96. \]

To perform the test, each of 20 students counted the number of yellow lollies
and the total number of lollies in a 48.1 gram packet.  Let's load the data:
```{r }
data <- read.table("candies.txt", header = TRUE)  # load the data
```

If each student made a test of $H_0\colon p=0.2$ at the 5\% level of
significance, what proportion of students rejected the null hypothesis?
```{r }
y <- data[, 1]
n <- data[, 2]
p <- y / n
z <- abs(p - 0.2) / sqrt(0.2 * 0.8 / n)
sum(z > 1.96) / length(z)  # proportion
which(z > 1.96)  # this shows *which* students rejected the null
```

If the null hypothesis were true, what proportion of students do you
expect to reject the null hypothesis at the 5\% level of significance?  
(Approximately $1 / 20 = 0.05$.)

For each of the 20 ratios in part 3, a 95\% confidence interval can be
constructed.  What proportion of these intervals contains $p = 0.2$?
```{r }
b1 <- p - 1.96 * sqrt(p * (1 - p) / n)
b2 <- p + 1.96 * sqrt(p * (1 - p) / n)
sum((b1 <= 0.2) & (0.2 <= b2)) / length(p)
```

If the 20 results are pooled do we reject $H_0\colon p = 0.2$?
```{r }
x <- sum(y)
N <- sum(n)
prop.test(x, N, p = 0.2, alternative = "two.sided")
```
We cannot reject the null hypothesis.  Therefore, from the data we do not have
enough evidence to reject the claim that 20\% of lollies in the packet are
yellow.


# Comparing two populations

Invadopodia are actin-rich protrusions of the plasma membrane that are
associated with degradation of the extracellular matrix in cancer invasiveness
and metastasis.  Dr Sloan treated one sample of cells using a drug which may
reduce the development of invadopodia, while another sample of cells received a
neutral treatment.  Using a microscope she counted the number of cells
developing invadopodia in each tissue portion.  In the treatment group she
found that 25 out of 351 cells developed invadopodia, while in the neutral
group 50 out of 389 cells developed invadopodia.

Carry out a two-sample z-test at the $\alpha = 0.05$ level of significance to
determine whether the treatment is effective in reducing the number of
invadopodia.  The following tests the null hypothesis $H_0\colon p_1 = p_2$
versus the two-sided alternative $H_1\colon p_1 \neq p_2$:
```{r }
x <- c( 25,  50)  # successes
n <- c(351, 389)  # sample sizes
prop.test(x, n, alternative = "two.sided")
```
The output above shows the value of the chi-square statistic defined as $Z^2$.
The observed value of the $Z$-statistic is $\vert z_{obs} \vert = \sqrt{6.0393}
= 2.4575$.  When $\alpha = 0.05$, the rejection region for this test is $\vert
z\vert > 1.96$.  Therefore, we reject $H_0$ and conclude that the treatment is
effective in reducing invadopodia development.


# Exercises

1. In the [Type I and Type II errors] section, the simulations did not give
answers that were particularly close to the true values.  Furthermore, the
confidence intervals for the simulations were quite wide, indicating that the
simulation estimates are not very precise.

    a) Explain why the simulations shown here are not very useful.
    ```{r }
    # Insert answer here...
    ```

    b) Improve the simulations and repeat them, showing that they indeed can
    give accurate and precise estimates.
    ```{r }
    # Insert answer here...
    ```

2. Refer to question 3 from the tutorial problems.

    a) Do the question.
    ```{r }
    # Insert answer here...
    ```

    b) Draw a power curve for a significance level of 0.05, for all possible
    values of $p$.
    ```{r }
    # Insert answer here...
    ```

3. Do question 4 from the tutorial problems.
```{r }
# Insert answer here...
```

4. A political party commissions an election poll, asking participants whether
they will vote for them rather than a rival party.  They will survey 900
people.  Let $p$ the true proportion of people who intend to vote for this
party.  Upon receiving the responses, they will carry out a hypothesis test of
$H_0\colon p = 0.5$ against $H_1\colon p \geqslant 0.5$, using a significance
level of 0.1.

    a) What is the test statistic and critical region?
    ```{r }
    # Insert answer here...
    ```

    b) What is the power when $p = 0.52$?
    ```{r }
    # Insert answer here...
    ```

    c) Draw a power curve for $0.5 \leqslant p \leqslant 0.6$.
    ```{r }
    # Insert answer here...
    ```

    d) The poll was run and 465 people responded in favour of the party.  Carry
    out the test and state a conclusion.
    ```{r }
    # Insert answer here...
    ```

    e) What can the party do to get a more conclusive result?
    ```{r }
    # Insert answer here...
    ```
