---
title: "Week 2 Lab, MAST20005/MAST90058"
author: "School of Mathematics and Statistics, University of Melbourne"
date: "Semester 2, 2022"
institute: "University of Melbourne"
fontsize: 11pt
classoption: a4paper
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
---

**Goals:** (i) Getting started with R and RStudio; (ii) Basic exploratory data
analyses; (iii) Basic graphics.


# R introduction and reference

This guide is available online at:  
<https://mathstats.pages.gitlab.unimelb.edu.au/r-reference/>

This online guide is much longer than the usual weekly lab notes. You are not
expected to absorb it completely in a single lab session.  We suggest you skim
it upon first reading and ensure you set aside **at least 15 minutes** to work
through the lab exercises at the end of your first lab class, referring back to
the online guide when required.


# Exercises

1. Calculate $\sum_{i=1}^{100} \ln(i)$.
```{r }
# Insert answer here...
```

2. Let $X \sim \mathrm{N}(1, 2)$.  What is $\mathbb{E}(X^2)$?  Approximate this
by simulating a large number of normal random variables and doing an appropriate calculation.
```{r }
# Insert answer here...
```

3. Use the help system to find out what the `qnorm()` function does. Explain
the result of running `qnorm(0.1)`.  What about `qnorm(0.1, lower.tail
= FALSE)`?  For what value of `a` will the command `qnorm(a)` return the same
value as the previous one?
```{r }
# Insert answer here...
```

4. Write a function `exp1pdf()` that calculates the pdf of an exponential
distribution with mean 1.  Compare the output of your function against the
in-built function that does the same calculation (`dexp()`).  Remember to check
it works for all inputs: for example, what is the correct value of
`exp1pdf(-1)`?
```{r }
# Insert answer here...
```

5. Do question 6(c) from the tutorial problems.
```{r }
# Insert answer here...
```

6. Do question 7(b) from the tutorial problems.
```{r }
# Insert answer here...
```

7. Do question 2(c) from the tutorial problems by simulation (rather than using
the Central Limit Theorem).
```{r }
# Insert answer here...
```
